---
title: "HW 04: Tidy Data and Joins"
author: "Julian Heavyside"
date: "10/10/2017"
output: 
  github_document:
    toc: true
---

### Part 1 Activity 2: Make a tibble with one row per year and columns for life expectancy for two or more countries

```{r, echo = FALSE}
#load relevant libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
knitr::opts_chunk$set(fig.width=6, fig.height=3)
```

I'm working with the two countries with the highest and lowest life expectancies (based on the highest LE ever reached in those countries) 

```{r}
# a convoluted way to find the countries with the largest and smallest LE
gapminder %>%
  group_by(country) %>% 
  summarise(top_LE = max(lifeExp)) %>% 
  arrange(top_LE) %>% 
  slice(c(1, n())) #this returns the first and last row of the summarised and ordered data
```

Japan's record high LE is almost twice that of Sierra Leone!


Now, the filtered dataset:

```{r}
siljap <- gapminder %>%
  select(year, country, lifeExp) %>% 
  filter(country == "Sierra Leone" | country == "Japan") %>% 
  spread(country, lifeExp) #create a column for each country
  
siljap %>% 
  knitr::kable(col.names = c("Year", "Sierra_Leone", "Japan"), align = "c")
```


I *hate* right aligned tables (`kable()`'s default alignment), so I was happy to discover the `align` parameter in the `kable()` function


Visualize the data (I found a cool [github package](https://github.com/baptiste/ggflags) created by [baptiste](https://github.com/baptiste) to plot images of country flags as data points. This is how you install and load the package from `babtiste`'s github repository

```{r}
suppressPackageStartupMessages(library(devtools))
install_github("baptiste/ggflags")
suppressPackageStartupMessages(library(ggflags))
```

The space between "Sierra" and "Leone" causes problems when plotting, and I used [this stackoverflow question](https://stackoverflow.com/questions/6081439/changing-column-names-of-a-data-frame) to remind myself how to change column names
```{r}
colnames(siljap)[which(names(siljap) == "Sierra Leone")] <- "Sierra_Leone"

siljap %>% 
  ggplot(aes(x = year)) +
  geom_flag(aes(y = Sierra_Leone, country = "sl"), size = 8) +
  geom_flag(aes(y = Japan, country = "jp"), size = 8) +
  xlab("Year") + 
  ylab("Life Expectancy")

```

Another way to compare the two countries is to plot them against eachother in a scatterplot. The current shape of the data is ideal for this. The two countries differ in LE so greatly that there is little point in plottin them with a line through the intercept to show a 1:1 ratio in LE. Japan is always ~2x greater than Sierra Leone

```{r}
siljap %>% 
  ggplot(aes(x=Sierra_Leone, y=Japan)) +
  geom_point() +
  geom_label(aes(label = year)) +
  xlab("Sierra Leone")

```

In both graphs we can see a blip in the LE for Sierra Leone around 1992. Civil war broke out in the country in 1991, and the government was overthrown after a coup d'état in 1992. This appears to be associated with a decrease in life expectancy during these few years. Read more [here](https://en.wikipedia.org/wiki/Sierra_Leone_Civil_War)



### Part 2 Join, merge, look up (Activity 1)

Mutating joins allow us to combine two tables, meaning that we can add variables from one table to another table, based on matching observations of variables that are shared between the two tables

First, I'll create a (small) dataset with one language and a popular food for speakers of that language, for a small selection of countries

```{r}

eatspeak<- data.frame(country = c("Canada", "Peru", "Italy", "France", "Norway", "Papua New Guinea", "New Zealand", "Afghanistan"),
                  language = c("Heiltsuk", "Quechua", "Neapolitan", "Provencal", "Sami", "Hiri Motu", "Maori", "Isken"),
                  food = c("Eulachon", "Chuño", "Pane cafone", "Truffle", "Cloudberry", "Kaukau", "Taewa tutaekuri", "Bolani"))

eatspeak

```

###`left_join`

Now I'll join some gapminder data to these countries. The result includes all the data from the 12 years, plus the two columns from `eatspeak`, for only the countries in `eatspeak` (filter out only one year to make the table nice and short)


```{r}
left_join(eatspeak, gapminder) %>% 
  filter(year == 2002) %>% 
  knitr::kable()

```

`gapminder` does not contain any data for Papua New Guinea, and the resulting table 

#`inner_join`








