---
title: "Homework 2"
date: '2017-09-26'
output: github_document
---

```{r}
suppressPackageStartupMessages(library(tidyverse))  # The tidyverse contains ggplot2!
suppressPackageStartupMessages(library(gapminder))
knitr::opts_chunk$set(fig.width=4, fig.height=3)
```

## Smell test the data

```{r}
str(gapminder)
```

The `gapminder` object is a dataframe. Its class is 'tbl_df'

There are 6 variables

There are 1704 observations

You can use the following functions to figure out what's in the `gapminder` dataset

```{r}
summary(gapminder)

typeof(gapminder)

class(gapminder)

str(gapminder)
```

Running `str(gapminder)` will return the type of each variable, as well as some other useful info

## Explore individual variables

Categorical var = `country`
Quantitative var = `gdpPercap`

Range in `gdpPercap`
```{r}
range(gapminder$gdpPercap)
```
Possible values for `country`

```{r}
unique(gapminder$country)
```

maximum gdp per capita for each country
```{r}
gapminder %>% 
  group_by(country) %>% 
  summarise(mingdppercap = max(gdpPercap))
```

change in gdp per capita over time, for all countries

```{r}
gapminder %>% 
  ggplot(aes(year, gdpPercap)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## Explore various plot types

gdp per capita across years, for each continent

```{r}
gapminder %>% 
  ggplot(aes(year, gdpPercap,
             colour = continent)) +
  geom_point() +
  geom_smooth(method = "lm")
```

looking just at Asia

```{r}
gapminder %>% 
  filter(continent == "Asia") %>% 
  ggplot(aes(year, gdpPercap,
             colour = country)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Evaluate the code below

```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))

```

Better to do it this way
```{r}
filter(gapminder, country == "Rwanda" | country == "Afghanistan")
```


