---
title: "hw03"
author: "Julian Heavyside"
date: "9/29/2017"
output: github_document
---

```{r}
suppressPackageStartupMessages(library(tidyverse))  # The tidyverse contains ggplot2!
suppressPackageStartupMessages(library(gapminder))
knitr::opts_chunk$set(fig.width=4, fig.height=3)
```

## Homework 3: Using dplyr and ggplot2 to manipulate and explore the gapminder dataset

# Homework Tasks

__Task 1__: Get the maximum and minimum of GDP per capita for all continents. I also want to know which country is the lowest for each continent

```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarise(minGDP = min(gdpPercap), maxGDP = max(gdpPercap)) %>% 
  knitr::kable() # make the table look prettier
```

__Task 2__: Look at the spread of GDP per capita within the continents.

```{r}

```

__Task 3__: Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.

```{r}

```

__Task 4__: How is life expectancy changing over time on different continents?

```{r}

```


__Task 5__: Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.


__Task 6__: Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged. This will generate interesting questions to follow up on in class.